---
title: "Lab | Introduction to R and Tidyverse"
output: html_notebook
---

1. **Install and load the Tidyverse**:
If you haven't already, install the tidyverse package and load it into your R session.

```{r}
install.packages("tidyverse")
library(tidyverse)
```

2.**Create a data frame:** 
Use the following code to create a sample data frame for this exercise.

```{r}
employees <- data.frame(
		  ID = 1:6,
		  Name = c("Alice", "Bob", "Charlie", "David", "Eve", "Frank"),
		  Age = c(25, 30, 35, 40, 45, 50),
		  Department = c("HR", "IT", "Finance", "IT", "HR", "Finance"),
		  Salary = c(50000, 60000, 70000, 80000, 55000, 75000)
		)
```
This data frame contains information about employees, including their ID, name, age, department, and salary.

### Task 1: Explore the Data
1. Print the employees data frame to the console.
```{r}
print(employees)
```

2. Use the str() function to inspect the structure of the data frame.
```{r}
str(employees)
```
3. Use the summary() function to get a summary of the data.
```{r}
summary(employees)
```
### Task 2: Basic Data Manipulation with dplyr

1. Filter rows: Create a new data frame that includes only employees who work in the "IT" department.
```{r}
it_employees <- employees %>% filter(Department == "IT")
print(it_employees)
```

2. Select columns: Create a new data frame that includes only the Name and Salary columns.
```{r}
name_salary_df<-employees %>% select(Name,Salary)
print(name_salary_df)
```

3. Add a new column: Add a new column called Bonus that calculates a 10% bonus for each employee based on their salary.
```{r}
mutated_df<- employees %>% mutate(Bonus=Salary*0.10)
print(mutated_df)
```

4. Sort rows: Sort the employees data frame by Salary in descending order.
```{r}
employees_sorted<-employees %>% arrange(desc(Salary))
print(employees_sorted)
```

5. Summarize data: Calculate the average salary for each department.
```{r}
average_salary_by_dept<- employees %>% 
group_by(Department) %>%
summarize(Average_Salary=mean(Salary))
print(average_salary_by_dept)
```
## Optional Exercises

### Extra 1: Advanced Data Manipulation

1. Group and summarize: Group the data by Department and calculate the total salary expenditure for each department.
```{r}
total_salary_exp_by_dept <- employees %>%
group_by(Department) %>%
summarize(Total_Salary = sum(Salary))
print(total_salary_exp_by_dept)
```

2. Filter and mutate: Create a new data frame that includes only employees older than 30 and adds a column called Experience that assumes each employee has Age - 25 years of experience.
```{r}
employees_30_above_df <- employees %>%
filter(Age>30) %>%
mutate(Experience=Age-25)
print(employees_30_above_df)
```
### Extra 2: Challenge

1. Combine operations: Create a new data frame that includes employees from the "HR" department, adds a Bonus column (10% of salary), and sorts the data by Bonus in descending order.
```{r}
HR_dept_df <- employees %>%
  filter(Department=="HR") %>%
  mutate(Bonus=Salary*0.10) %>%
  arrange(desc(Bonus))
print(HR_dept_df)
```

2. Visualize data: Use ggplot2 to create a bar plot showing the total salary expenditure by department.
```{r}
library(ggplot2)
total_salary_by_dept <- employees %>%
  group_by(Department) %>%
  summarize(TotalSalary =sum(Salary))
            
  ggplot(total_salary_by_dept, aes(x=Department, y=TotalSalary, fill=Department))+
    geom_bar(stat='identity') +
    labs(title='Total Salary Expenditure by Department',
         x='Departmnent',
         y='Total Salary Expenditure') +
    theme_minimal()
```


